Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-08-08 volunteer-clouds-ieee-ci-si.tex
Adds description of results

\begin{abstract}
Volunteer computing is used routinely in science and industry, but it
still has a long way to go if it is going to become a mainstream technology in the
same way the cloud technologies in general have. The lack of general
purpose and open source tools 
with analytics built in could be one of the reasons why this is so, as
well as the lack of proofs of concept. That is why in this paper we
present NodIO, a client-server architecture for distributed
evolutionary algorithms, written entirely in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser. The architecture, along with some initial results on the
expected performance and scaling 
behavior is presented. We also use the algorithm data to 
find out what kind of patterns the user behavior follow and try to fit
them to some statistical distribution. 
\end{abstract}

@@

There is, then, space for systems that do not have any of these 
assumptions, such as the one presented in this paper, which will be
provisionally named {\sf NodIO}. NodIO is first an application that
runs in two tiers, a server and a client, the latter  can be possibly
run on a  % [Pedro] Both the server and the client in browsers?
          % different browsers as I understand? [JJ]  No, server in
          % server, client in browser, although there's nothing that
          % can't be done in a browser. 
browser. As such, it can be run in any computer that loads the web
page where the client is embedded, including many over which%[Mario]Quitar el possibly?   [JJ] Done
the scientist has no control.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{img/baseline-times.png}
\caption{Comparison of time to solution for the baseline system, using a {\tt
    node.js} client, for different population sizes. Only
runs in which solution has been found are used for this graph.}
\label{fig:baseline}
\end{figure}
First we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using NodEO \cite{figshare:nodeo}, the basic JavaScript library we have used as base to built {\sf NodIO}. This experiment tries to 
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$ and a population size equal 512. The algorithm was run until the solution (a string with all
ones) was found or five million evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with success rate upgraded to 100\% and average duration $3.46$
seconds. Results for these two experiments 
are shown in Figure \ref{fig:baseline}.

@@

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented 
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two NodEO instances per browser. The sequence diagram shows the 
interaction between processess in a time sequence. In this system there are
two kinds of asynchronus messages: HTTP Requests shown as bold arrows between 
the client and the server where PUT and GET HTTP Methods are represented by 
red and blue arrows respectively. There are also messages between workers and 
the main thread in locally in the client, this are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em nodejs} This is the server side process, implented in the nodejs 
framework. Web request handlers are implemented using the Express web 
framework. For keeping a log of the execution the Winston library is used, and 
finally the NodeIO framework is responsible for keeping the shared pool.
\item {\em Client main thread}  This process handles the user interaction
using the JQuery framwork. The Chart.js library is used to display a chart of
the best fitness in the population over time. The main script also creates 
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have 
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronus HTTP requests to the server are implemented using the XHR object. 
\end{itemize}

The sequence of events are:
\begin{enumerate}
\item First the NodIO process is started in nodejs. As part of the 
initialization procedure the number of experiment is initialised and a log 
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP 
request and the page is loaded.
\item The main script renders the page and creates the worker instance. In 
order to create a dedicated worker instance another GET request is sent to the 
server in order to recieve the worker implementation script. 
Each worker instance executes one or more evolutionary algorithms 
{\em islands}. Each starts with random population and it is assigned a 
universally unique  identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.  
The initial paramaters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following with the previously 
specified parameters. Once the population has been initialised, the EA 
generation loop is started. After n generations the state of the population 
is evaluated. First the chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not a global best,
the EA loop must continue, but first a chromosome is randomly selected from 
the shared pupulation. Also the current generation number and the best 
chromosome is sent to the client main thread in order to update the DOM to 
display the current state of the island (described next). These messages are
all asynchronous so even before a message is recieved by target the EA loop of
the worker could continue. If the best fitness is a global best the best 
chromosme is sent to the server and the loop ends. A message is sent to the 
main thread indicating that the best individual has been found.
\item The main thread recieves messages from workers as events handled by 
a callback function. When an iteration message is recived ,the page is 
updated by displaying a dinamic plot of the geneartion number and best 
fitness, and a representation of the chromosme is also generated.   
\item When a global best is recieved from an island the current experiment 
ends, the experiment number is incremented and the population array is reset.  
\item When the main thread recieves a message the a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the cost, only the parameters and population is reset and a new UIID 
assigned.
\end{enumerate}
%
\begin{figure}[!t]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}  
\label{fig:system:w2}
\end{figure}
%

\begin{table*}
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline 
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline 
\end{tabular}
\end{center}
\end{table*}
%
%---------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

@@

This work has been supported in part by TIN2011-28627-C04-02 and. 
TIN2014-56494-C4-3-P (Spanish Ministry of Economy and Competitivity),
SPIP2014-01437 (Direcci{\'o}n General de Tr{\'a}fico) and PYR-2014-17
GENIL project (CEI-BIOTIC Granada). We would also like to thank the
anonymous reviewers of previous versions of this paper who have really helped us to improve
this paper (and our work) with their suggestions. We are also grateful
to Anna Saéz de Tejada for her help with the data processing scripts.

>>>>

\begin{abstract}
Volunteer computing is used routinely in science and industry, but it
still has a long way to go if it is going to become a mainstream technology in the
same way the cloud technologies in general have. The lack of general
purpose and open source tools 
with analytics built in could be one of the reasons why this is so, as
well as the lack of proofs of concept. That is why in this paper we
present {\sf NodIO}, a client-server architecture for distributed
evolutionary algorithms, written entirely in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser. The architecture, along with some initial results on the
expected performance and scaling 
behavior is presented. We also use the algorithm data to 
find out what kind of patterns the user behavior follow and try to fit
them to some statistical distribution. 
\end{abstract}

@@

There is, then, space for systems that do not have any of these 
assumptions, such as the one presented in this paper, which will be
provisionally named {\sf NodIO}. {\sf NodIO} is first an application that
runs in two tiers, a server and a client, the latter  can be possibly
run on a  % [Pedro] Both the server and the client in browsers?
          % different browsers as I understand? [JJ]  No, server in
          % server, client in browser, although there's nothing that
          % can't be done in a browser. 
browser. As such, it can be run in any computer that loads the web
page where the client is embedded, including many over which%[Mario]Quitar el possibly?   [JJ] Done
the scientist has no control.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{img/baseline-times.png}
\caption{Comparison of time to solution for the baseline system, using a {\tt
    node.js} client, for different population sizes. Only
runs in which solution has been found are used for this graph.}
\label{fig:baseline}
\end{figure}
First we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using NodEO \cite{nodeo2014}, the basic JavaScript library we have used as base to built {\sf NodIO}. This experiment tries to 
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$ and a population size equal 512. The algorithm was run until the solution (a string with all
ones) was found or five million evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with success rate upgraded to 100\% and average duration $3.46$
seconds. Results for these two experiments 
are shown in Figure \ref{fig:baseline}.

@@

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented 
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two NodEO instances per browser. The sequence diagram shows the 
interaction between processess in a time sequence. In this system there are
two kinds of asynchronus messages: HTTP Requests shown as bold arrows between 
the client and the server where PUT and GET HTTP Methods are represented by 
red and blue arrows respectively. There are also messages between workers and 
the main thread in locally in the client, this are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em nodejs} This is the server side process, implented in the nodejs 
framework. Web request handlers are implemented using the Express web 
framework. For keeping a log of the execution the Winston library is used, and 
finally the NodeIO framework is responsible for keeping the shared pool.
\item {\em Client main thread}  This process handles the user interaction
using the JQuery framework. The Chart.js library is used to display a chart of
the best fitness in the population over time. The main script also creates 
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have 
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronus HTTP requests to the server are implemented using the XHR object. 
\end{itemize}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}  
\label{fig:system:w2}
\end{figure}

The sequence of events is as follows:
\begin{enumerate}
\item First the {\sf NodIO} process is started in nodejs. As part of the 
initialization procedure the number of experiment is initialised and a log 
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP 
request and the page is loaded.
\item The main script renders the page and creates the worker instance. In 
order to create a dedicated worker instance another GET request is sent to the 
server in order to recieve the worker implementation script. 
Each worker instance executes one or more evolutionary algorithms 
{\em islands}. Each starts with random population and it is assigned a 
universally unique  identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.  
The initial paramaters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following with the previously 
specified parameters. Once the population has been initialised, the EA 
generation loop is started. After n generations the state of the population 
is evaluated. First the chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not a global best,
the EA loop must continue, but first a chromosome is randomly selected from 
the shared pupulation. Also the current generation number and the best 
chromosome is sent to the client main thread in order to update the DOM to 
display the current state of the island (described next). These messages are
all asynchronous so even before a message is recieved by target the EA loop of
the worker could continue. If the best fitness is a global best the best 
chromosme is sent to the server and the loop ends. A message is sent to the 
main thread indicating that the best individual has been found.
\item The main thread recieves messages from workers as events handled by 
a callback function. When an iteration message is recived ,the page is 
updated by displaying a dinamic plot of the geneartion number and best 
fitness, and a representation of the chromosme is also generated.   
\item When a global best is recieved from an island the current experiment 
ends, the experiment number is incremented and the population array is reset.  
\item When the main thread recieves a message the a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the cost, only the parameters and population is reset and a new UIID 
assigned.
\end{enumerate}
%
\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline 
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline 
\end{tabular}
\end{center}
\end{table*}
%

The experiments were run in the same way as before, using several
Twitter announcements throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results are shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, being considered every worker a different IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiment take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that is,
most of the times, better than we would using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.

However, it is interesting to note why this is so, and the first hint
is the inter-experiment correlation between the number of IPs in
succesive experiments. While before it was an unremarkable 10\%, it is
now more than 43\%, making the numbef of IPs in contiguous experiments
highly correlated. The main reason for this is the setup in {\sf
  NodIO-W$^2$}, which restarts an island after a solution has been
found and also is running even if the tab is not on the
foreground. This means that volunteers can leave the experiment
running for as long as they want and they will be contributing to
experiment after experiment, unlike the previous version, where they
stopped contributing after the solution was found. This is also
reflected in the number of PUTs, which is the number of generations
divided by 100; more than 50\% of the volunteers run for more than
4000 generations. All things considered, this means than there are
many more computers running {\em simultaneously}, leading to this
almost 5-fold increase in running time. 
%---------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

@@

This work has been supported in part by TIN2011-28627-C04-02 and. 
TIN2014-56494-C4-3-P (Spanish Ministry of Economy and Competitivity),
SPIP2014-01437 (Direcci{\'o}n General de Tr{\'a}fico) and PYR-2014-17
GENIL project (CEI-BIOTIC Granada). We would also like to thank the
anonymous reviewers of previous versions of this paper who have really helped us to improve
this paper (and our work) with their suggestions. We are also grateful
to Anna S\'aez de Tejada for her help with the data processing scripts.
